Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-16 latex/main.tex
Pango Xs to Pango X lineages
or just X lineages

Closes  #652

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e. the Pango Xs) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango Xs,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

@@

\subsection{Additional analyses of the Pango Xs}
\label{sec:supplement_indepth_pango_x_lineages}

@@

\subsection*{Multiple breakpoints}
Seven of the Pango Xs present in the Viridian v04 dataset have more than one breakpoint,
with the supporting evidence documented in the associated Pango designation issues.
Five of these seven Pango Xs were detected as recombinants by sc2ts.
Three of them (XAY, XBC, XBT) had more than one breakpoint,
but were not added to the ARG due to high HMM cost and
a lack of sample support to be incorporated as retrospective sample groups.
However, the other two (XAC and XAK) had only one breakpoint.

@@

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357
Evidence of the second breakpoint reported in its designation issue relies on
the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
The XAC samples share a recombination node with the samples of four other Pango Xs
which do not have this deletion (Figure 2C).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Classification of ``robust'' and artifactual recombination events.
Scatterplot of recombination events by the net number of supporting loci
for the left parent (x-axis) and for the right parent (y-axis). The shaded
regions highlight artifactual recombination events, which have fewer than four
net supporting loci on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing. The pie charts show
breakdowns of the recombination events by primer scheme per quadrant.
Recombination events associated with the origins of Pango Xs are labeled.
}
\label{fig:recombinant_qc}
\end{figure}

>>>>

We used Pango lineage designations to select representative samples
for the majority of the 2026 Pango lineages in the ARG.
We first identify an ``origination node" for a focal Pango lineage,
defined as the oldest internal node in the ARG
whose descendants in any tree include more than 50\% of the Pango samples of the focal type,
and where the internal node is itself designated as the focal Pango.
The representative sample is then chosen as the oldest sample descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We ``simplify''\cite{Kelleher2018efficient,Wong2024} 
the ARG to include only the representative samples,
resulting in an intermediate ARG of 1867 samples,
then additionally remove internal samples,
known recombinants (i.e.\ the Pango X lineages) and three ``singleton recombinant
samples''
in which a single Pango lineage appears immediately below a recombination node.
This results in a simplified ARG of 1667 samples, each with a different Pango designation.

@@

For CovRecomb, we summarized the detection results for the Pango X lineages as follows.
First, we obtained the "independent" recombination events
identified by Li et al.\cite{Li2024CovRecomb} (see Table S4 of the study)
which correspond to the Pango X lineages (if available).
When multiple events corresponded to a Pango X lineage,
we took the results of the event which had the highest number of epidemic recombinants.
For other Pango X lineages, we downloaded the full list of putative recombinants
identified in the study
(https://github.com/wuaipinglab/CovRecomb/blob/main/CovRecomb-Global-Version/putative\_recombinants/putative\%20recombinants.csv)
For each of these Pango X lineages,
we took the most commonly observed combination of
the left parent, the right parent, and the breakpoint interval.
The breakpoint intervals are defined as the left and right recombination-informative sites.
We adjusted the breakpoint intervals by adding one to the right coordinate
so that they are right-exclusive.

@@

\subsection{Additional analyses of the Pango X lineages}
\label{sec:supplement_indepth_pango_x_lineages}

@@

\subsection*{Multiple breakpoints}
Seven of the Pango X lineages present in the Viridian v04 dataset have more than one breakpoint,
with the supporting evidence documented in the associated Pango designation issues.
Five of these seven X lineages were detected as recombinants by sc2ts.
Three of them (XAY, XBC, XBT) had more than one breakpoint,
but were not added to the ARG due to high HMM cost and
a lack of sample support to be incorporated as retrospective sample groups.
However, the other two (XAC and XAK) had only one breakpoint.

@@

\subsubsection*{XAC}
https://github.com/jeromekelleher/sc2ts-paper/issues/357
Evidence of the second breakpoint reported in its designation issue relies on
the presence of a large deletion inherited from BA.2*, which is currently treated as missing data by sc2ts.
The XAC samples share a recombination node with the samples of four other 
X lineages
which do not have this deletion (Figure 2C).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_qc}
\caption{
Classification of ``robust'' and artifactual recombination events.
Scatterplot of recombination events by the net number of supporting loci
for the left parent (x-axis) and for the right parent (y-axis). The shaded
regions highlight artifactual recombination events, which have fewer than four
net supporting loci on one side or both sides of the suggested breakpoint.
Colors indicate the primer scheme used for sequencing. The pie charts show
breakdowns of the recombination events by primer scheme per quadrant.
Recombination events associated with the origins of Pango X lineages are labeled.
}
\label{fig:recombinant_qc}
\end{figure}
